---
title: "How Does Age Affect Soccer Player Performance?"
---

## Project Overview

In professional soccer, understanding how an athlete's age relates to on-field production is vital for roster construction, tactical planning, and player valuation. This project examines the relationship between player age and offensive output across Europe's top five domestic leagues (Premier League, La Liga, Serie A, Bundesliga, and Ligue 1) during the 2025–2026 season.

Our primary performance metric is **goal contributions per 90 minutes**, calculated as:

$$\text{Goal Contributions per 90} = \frac{\text{Goals} + \text{Assists}}{\text{90s Played}}$$

Normalizing per 90 minutes allows us to compare players who logged different total amounts of playing time across the season.

```{r}
#| label: setup-homepage
#| message: false
#| warning: false
#| echo: false

library(readr)
library(dplyr)
library(ggplot2)
library(plotly)

players_raw <- read_csv("data/players_data_light-2025_2026.csv", show_col_types = FALSE)

players <- players_raw |>
  filter(!is.na(Age), !is.na(Gls), !is.na(Ast), !is.na(`90s`), `90s` > 0) |>
  mutate(ga_per_90 = (Gls + Ast) / `90s`)

n_obs <- nrow(players)
```

## Age vs. Goal Contributions

**Total observations analyzed:** `r format(n_obs, big.mark = ",")` players.

```{r}
#| label: fig-age-performance
#| fig-cap: "Scatterplot of player age versus goal contributions per 90 minutes across Europe's top five leagues, with a LOESS smooth trend line."
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| fig-height: 5.5

p <- ggplot(players, aes(x = Age, y = ga_per_90)) +
  geom_point(
    aes(text = paste0(
      "<b>", Player, "</b>",
      "<br>Age: ", Age,
      "<br>Squad: ", Squad,
      "<br>League: ", sub("^[a-z]{2} ", "", Comp),
      "<br>G+A per 90: ", round(ga_per_90, 2)
    )),
    alpha = 0.3, color = "#1d3557", size = 1.8
  ) +
  geom_smooth(method = "loess", color = "#e63946", se = TRUE, linewidth = 1.2) +
  coord_cartesian(ylim = c(0, 2.5)) +
  scale_x_continuous(breaks = seq(15, 45, by = 5)) +
  labs(
    title = "Player Age and Goal Contributions per 90 Minutes",
    x = "Age (Years)",
    y = "Goal Contributions per 90 (Gls + Ast / 90s)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )

gp <- ggplotly(p, tooltip = "text")
if (length(gp$x$data) >= 2) gp$x$data[[2]]$hoverinfo <- "none"
if (length(gp$x$data) >= 3) gp$x$data[[3]]$hoverinfo <- "none"

gp
```

### Interpreting the Pattern

The scatterplot above displays all `r format(n_obs, big.mark = ",")` player records from the 2025–2026 season. The LOESS smoothing curve indicates a modest peak in average goal contributions in the mid-to-late twenties, followed by a slight decline into the thirties. 

**Important note on correlation vs. causation:** This graphic shows an observational association, not a causal relationship. Age itself does not determine a player's goal contribution rate. Factors such as tactical position, team strength, survivor bias (only elite older players remain in top leagues), and role specialization heavily influence these numbers.

### Data Source
Data was retrieved from Kaggle: [Football Players Statistics (2025–2026)](https://www.kaggle.com/datasets/hubertsidorowicz/football-players-stats-2025-2026) by Hubert Sidorowicz, sourced from FBref.

