---
toc: false
page-layout: full
---

```{=html}
<div class="home-hero">
  <div class="home-hero__media" aria-hidden="true"></div>
  <div class="home-hero__overlay"></div>
  <div class="home-hero__content">
    <p class="home-hero__eyebrow">Age, performance, and attacking output</p>
    <h1>How Does Age Affect Soccer Player Performance?</h1>
    <p class="home-hero__intro">Across Europe's top leagues, we compare the age curve of attacking production to show how experience, role, and survival bias shape the numbers behind elite players.</p>
    <div class="home-hero__actions">
      <a class="home-hero__button home-hero__button--primary" href="#age-vs-goal-contributions">Explore the Data</a>
    </div>
  </div>
</div>
```

```{r}
#| label: setup-homepage
#| message: false
#| warning: false
#| echo: false

library(readr)
library(dplyr)
library(ggplot2)
library(DT)
library(htmltools)
library(stringr)

players_raw <- read_csv("data/players_data_light-2025_2026.csv", show_col_types = FALSE)

players <- players_raw |>
  filter(!is.na(Age), !is.na(Gls), !is.na(Ast), !is.na(`90s`), `90s` > 0) |>
  mutate(ga_per_90 = (Gls + Ast) / `90s`)

n_obs <- nrow(players)

players_table <- players_raw |>
  filter(!is.na(Age), !is.na(Gls), !is.na(Ast), !is.na(`90s`), `90s` > 0) |>
  mutate(
    League = str_replace(Comp, "^[a-z]{2} ", ""),
    Position = str_replace_all(Pos, ",", " / "),
    `Goal Contributions` = Gls + Ast,
    `Goals / 90` = round(Gls / `90s`, 2),
    `Assists / 90` = round(Ast / `90s`, 2),
    `Goal Contributions / 90` = round((Gls + Ast) / `90s`, 2)
  ) |>
  select(
    Player,
    Age,
    Squad,
    League,
    Position,
    Gls,
    Ast,
    `Goal Contributions`,
    `Goals / 90`,
    `Assists / 90`,
    `Goal Contributions / 90`
  ) |>
  rename(Team = Squad, Goals = Gls, Assists = Ast) |>
  arrange(desc(`Goal Contributions`), desc(Age))

team_choices <- sort(unique(players_table$Team))
league_choices <- sort(unique(players_table$League))
position_choices <- sort(unique(players_table$Position))
```

## Age vs. Goal Contributions {#age-vs-goal-contributions}

**Total observations analyzed:** `r format(n_obs, big.mark = ",")` players.

```{r}
#| label: fig-age-performance
#| fig-cap: "Scatterplot of player age versus goal contributions per 90 minutes across Europe's top five leagues, with a LOESS smooth trend line."
#| warning: false
#| message: false
#| echo: false
#| fig-width: 9
#| fig-height: 5.5

p <- ggplot(players, aes(x = Age, y = ga_per_90)) +
  geom_point(alpha = 0.34, color = "#26744f", size = 1.8) +
  geom_smooth(method = "loess", color = "#e87854", se = TRUE, linewidth = 1.2) +
  coord_cartesian(ylim = c(0, 2.5)) +
  scale_x_continuous(breaks = seq(15, 45, by = 5)) +
  labs(
    title = "Player Age and Goal Contributions per 90 Minutes",
    x = "Age (Years)",
    y = "Goal Contributions per 90 (Gls + Ast / 90s)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#10231d"),
    axis.title = element_text(color = "#263a30"),
    axis.text = element_text(color = "#586b61"),
    panel.grid.major = element_line(color = "#d5dfd7"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f2f5ef", color = NA),
    plot.background = element_rect(fill = "#f2f5ef", color = NA)
  )

p
```

### Interpreting the Pattern

The scatterplot above displays all `r format(n_obs, big.mark = ",")` player records from the 2025–2026 season. The LOESS smoothing curve indicates a modest peak in average goal contributions in the mid-to-late twenties, followed by a slight decline into the thirties.

**Important note on correlation vs. causation:** This graphic shows an observational association, not a causal relationship. Age itself does not determine a player's goal contribution rate. Factors such as tactical position, team strength, survivor bias (only elite older players remain in top leagues), and role specialization heavily influence these numbers.

## Project Overview

:::: {.overview-grid}

::: {.overview-copy}
In professional soccer, age and output shape recruitment, squad planning, and player valuation. This report tracks attacking production across the Premier League, La Liga, Serie A, Bundesliga, and Ligue 1 during the 2025-2026 season.

We compare players using **goal contributions per 90 minutes**, which adjusts goals and assists for playing time.
:::

::: {.metric-panel}
<div class="metric-label">RATE MODEL / PER 90</div>

$$
	ext{Goal Contributions per 90} = \frac{\text{Goals} + \text{Assists}}{\text{90s Played}}
$$

<div class="metric-note">GOALS + ASSISTS / 90 MINUTES PLAYED</div>
:::

::::

## Players and Goal Contributions

Goal contributions are defined as goals plus assists. Use the controls below to search by player name, narrow the age range, and filter by team, league, or position. The default sort is highest total goal contributions.

```{r}
#| echo: false
#| warning: false
#| message: false

player_table_controls <- tagList(
  div(
    class = "player-table-controls",
    div(
      class = "player-table-search",
      tags$label("Search players", `for` = "player-search"),
      tags$input(
        id = "player-search",
        type = "search",
        placeholder = "Search by name",
        class = "player-table-input"
      )
    ),
    div(
      class = "player-table-filter-group",
      tags$label("Min age", `for` = "age-min"),
      tags$input(id = "age-min", type = "number", min = 15, max = 45, step = 1, placeholder = "Min", class = "player-table-input player-table-number"),
      tags$label("Max age", `for` = "age-max"),
      tags$input(id = "age-max", type = "number", min = 15, max = 45, step = 1, placeholder = "Max", class = "player-table-input player-table-number"),
      tags$label("Team", `for` = "team-filter"),
      tags$select(
        id = "team-filter",
        class = "player-table-input",
        tags$option(value = "", "All teams"),
        lapply(team_choices, function(x) tags$option(value = x, x))
      ),
      tags$label("League", `for` = "league-filter"),
      tags$select(
        id = "league-filter",
        class = "player-table-input",
        tags$option(value = "", "All leagues"),
        lapply(league_choices, function(x) tags$option(value = x, x))
      ),
      tags$label("Position", `for` = "position-filter"),
      tags$select(
        id = "position-filter",
        class = "player-table-input",
        tags$option(value = "", "All positions"),
        lapply(position_choices, function(x) tags$option(value = x, x))
      ),
      tags$button(id = "reset-player-filters", type = "button", class = "player-table-reset", "Reset filters")
    ),
    div(
      class = "player-table-summary",
      tags$span(id = "matching-player-count", "0 matching players")
    )
  ),
  div(
    id = "player-table-empty",
    class = "player-table-empty",
    hidden = NA,
    tags$p("No players match the selected filters."),
    tags$button(id = "empty-reset-player-filters", type = "button", class = "player-table-reset compact", "Reset filters")
  ),
  withTags(
    table(
      id = "player-goal-contributions-table",
      class = "cell-border stripe compact hover",
      thead(
        tr(
          lapply(c("Player", "Age", "Team", "League", "Position", "Goals", "Assists", "Goal Contributions", "Goals / 90", "Assists / 90", "Goal Contributions / 90"), th)
        )
      ),
      tbody(
        lapply(seq_len(nrow(players_table)), function(i) {
          row <- players_table[i, ]
          tr(
            hidden = if (i > 25) NA else NULL,
            td(row$Player),
            td(row$Age),
            td(row$Team),
            td(row$League),
            td(row$Position),
            td(format(row$Goals, big.mark = ",", scientific = FALSE)),
            td(format(row$Assists, big.mark = ",", scientific = FALSE)),
            td(format(row$`Goal Contributions`, big.mark = ",", scientific = FALSE)),
            td(sprintf("%.2f", row$`Goals / 90`)),
            td(sprintf("%.2f", row$`Assists / 90`)),
            td(sprintf("%.2f", row$`Goal Contributions / 90`))
          )
        })
      )
    )
  ),
  div(
    id = "player-table-pagination",
    class = "player-table-pagination",
    hidden = NA,
    tags$span(id = "player-table-page-summary"),
    div(
      class = "player-table-page-actions",
      tags$button(id = "player-table-prev", type = "button", class = "player-table-reset compact", disabled = NA, "Previous"),
      tags$button(id = "player-table-next", type = "button", class = "player-table-reset compact", "Next")
    )
  ),
  tags$script(HTML("(function() {
    const table = document.getElementById('player-goal-contributions-table');
    const rows = Array.from(table.querySelectorAll('tbody tr'));
    const searchInput = document.getElementById('player-search');
    const ageMin = document.getElementById('age-min');
    const ageMax = document.getElementById('age-max');
    const teamFilter = document.getElementById('team-filter');
    const leagueFilter = document.getElementById('league-filter');
    const positionFilter = document.getElementById('position-filter');
    const resetBtn = document.getElementById('reset-player-filters');
    const emptyState = document.getElementById('player-table-empty');
    const countEl = document.getElementById('matching-player-count');
    const emptyResetBtn = document.getElementById('empty-reset-player-filters');
    const pagination = document.getElementById('player-table-pagination');
    const pageSummary = document.getElementById('player-table-page-summary');
    const previousBtn = document.getElementById('player-table-prev');
    const nextBtn = document.getElementById('player-table-next');
    const pageSize = 25;
    let currentPage = 0;

    function refreshTable(resetPage) {
      if (resetPage) currentPage = 0;
      const query = (searchInput.value || '').trim().toLowerCase();
      const minAge = ageMin.value === '' ? null : Number(ageMin.value);
      const maxAge = ageMax.value === '' ? null : Number(ageMax.value);
      const teamValue = (teamFilter.value || '').trim();
      const leagueValue = (leagueFilter.value || '').trim();
      const positionValue = (positionFilter.value || '').trim();
      const matchingRows = [];

      rows.forEach(function(row) {
        const cells = row.children;
        const playerName = (cells[0] ? cells[0].textContent : '').trim().toLowerCase();
        const age = Number((cells[1] ? cells[1].textContent : '').trim()) || 0;
        const team = (cells[2] ? cells[2].textContent : '').trim();
        const league = (cells[3] ? cells[3].textContent : '').trim();
        const position = (cells[4] ? cells[4].textContent : '').trim();

        const matchesQuery = !query || playerName.includes(query);
        const matchesMin = minAge === null || age >= minAge;
        const matchesMax = maxAge === null || age <= maxAge;
        const matchesTeam = !teamValue || team === teamValue;
        const matchesLeague = !leagueValue || league === leagueValue;
        const matchesPosition = !positionValue || position === positionValue;

        const showRow = matchesQuery && matchesMin && matchesMax && matchesTeam && matchesLeague && matchesPosition;
        row.hidden = true;
        if (showRow) matchingRows.push(row);
      });

      const count = matchingRows.length;
      countEl.textContent = count + ' matching player' + (count === 1 ? '' : 's');
      const hasMatches = count > 0;
      const pageCount = Math.ceil(count / pageSize);
      currentPage = Math.min(currentPage, Math.max(0, pageCount - 1));
      const firstIndex = currentPage * pageSize;
      matchingRows.slice(firstIndex, firstIndex + pageSize).forEach(function(row) {
        row.hidden = false;
      });

      pageSummary.textContent = hasMatches
        ? 'Showing ' + (firstIndex + 1) + '–' + Math.min(firstIndex + pageSize, count) + ' of ' + count
        : 'No players to show';
      previousBtn.disabled = currentPage === 0;
      nextBtn.disabled = currentPage >= pageCount - 1;
      pagination.hidden = pageCount <= 1;
      emptyState.hidden = hasMatches;
      table.classList.toggle('table-empty', !hasMatches);
    }

    function resetFilters() {
      searchInput.value = '';
      ageMin.value = '';
      ageMax.value = '';
      teamFilter.value = '';
      leagueFilter.value = '';
      positionFilter.value = '';
      refreshTable(true);
    }

    [searchInput, ageMin, ageMax, teamFilter, leagueFilter, positionFilter].forEach(function(el) {
      el.addEventListener('input', function() { refreshTable(true); });
      el.addEventListener('change', function() { refreshTable(true); });
    });

    resetBtn.addEventListener('click', resetFilters);
    emptyResetBtn.addEventListener('click', resetFilters);
    previousBtn.addEventListener('click', function() {
      currentPage -= 1;
      refreshTable(false);
    });
    nextBtn.addEventListener('click', function() {
      currentPage += 1;
      refreshTable(false);
    });
    refreshTable(true);
  })();"))
)

player_table_controls
```

### Data Source
Data was retrieved from Kaggle: [Football Players Statistics (2025–2026)](https://www.kaggle.com/datasets/hubertsidorowicz/football-players-stats-2025-2026) by Hubert Sidorowicz, sourced from FBref.

