---
title: "When Does Performance Peak?"
subtitle: "How age and playing position shape goal contributions across Europe's top five leagues."
---

<div class="story-page">

This 2025–2026 snapshot tracks goals plus assists per 90 minutes across five age bands. The overall distribution and position-level comparisons show why one league-wide “peak age” can be misleading.

```{r}
#| label: setup-story
#| echo: false
#| message: false
#| warning: false

library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(htmltools)

players_raw <- read_csv("data/players_data_light-2025_2026.csv", show_col_types = FALSE)

qualified_players <- players_raw |>
  filter(!is.na(Age), !is.na(Gls), !is.na(Ast), !is.na(`90s`), `90s` > 0, Min >= 450) |>
  mutate(
    ga_per_90 = (Gls + Ast) / `90s`,
    age_group = case_when(
      Age <= 20 ~ "17–20",
      Age <= 24 ~ "21–24",
      Age <= 28 ~ "25–28",
      Age <= 32 ~ "29–32",
      TRUE ~ "33+"
    ),
    age_group = factor(age_group, levels = c("17–20", "21–24", "25–28", "29–32", "33+")),
    pos_label = case_when(
      grepl("^FW", Pos) ~ "Forwards",
      grepl("^MF", Pos) ~ "Midfielders",
      grepl("^DF", Pos) ~ "Defenders",
      grepl("^GK", Pos) ~ "Goalkeepers",
      TRUE ~ "Other"
    ),
    pos_label = factor(pos_label, levels = c("Forwards", "Midfielders", "Defenders", "Goalkeepers"))
  )

n_story <- nrow(qualified_players)
league_count <- n_distinct(qualified_players$Comp)
age_summary <- qualified_players |>
  group_by(age_group) |>
  summarise(sample_size = n(), mean_ga = mean(ga_per_90), .groups = "drop")

qualified_players <- qualified_players |>
  left_join(age_summary |> select(age_group, sample_size), by = "age_group") |>
  mutate(
    box_hover = paste0(
      "Age group: ", age_group,
      "<br>Metric: ", sprintf("%.4f", ga_per_90), " G+A per 90",
      "<br>Sample size: ", sample_size
    )
  )

position_summary <- qualified_players |>
  filter(!is.na(pos_label)) |>
  group_by(pos_label, age_group) |>
  summarise(
    sample_size = n(),
    mean_ga = mean(ga_per_90),
    .groups = "drop"
  ) |>
  mutate(
    hover = paste0(
      "Position: ", pos_label,
      "<br>Age group: ", age_group,
      "<br>Mean: ", sprintf("%.4f", mean_ga), " G+A per 90",
      "<br>Sample size: ", sample_size
    ),
    mean_label = sprintf("%.4f", mean_ga)
  )

position_colors <- c(
  "Forwards" = "#e87854",
  "Midfielders" = "#26744f",
  "Defenders" = "#a4d353",
  "Goalkeepers" = "#68766e"
)

make_position_plot <- function(data, compare = FALSE) {
  chart <- ggplot(
    data,
    aes(x = .data$age_group, y = .data$mean_ga, fill = .data$pos_label, text = .data$hover)
  ) +
    geom_col(width = 0.68, show.legend = FALSE) +
    geom_text(aes(label = .data$mean_label), vjust = -0.45, size = 3.5, show.legend = FALSE) +
    scale_fill_manual(values = position_colors, drop = FALSE) +
    scale_y_continuous(
      labels = scales::label_number(accuracy = 0.0001),
      expand = expansion(mult = c(0, 0.2))
    ) +
    labs(x = "Age group", y = "Mean G+A per 90") +
    theme_minimal(base_size = 13) +
    theme(
      axis.title = element_text(color = "#263a30", face = "bold"),
      axis.text = element_text(color = "#35483e", size = 11),
      panel.grid.major = element_line(color = "#d5dfd7"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "#ffffff", color = NA),
      plot.background = element_rect(fill = "#ffffff", color = NA),
      strip.text = element_text(face = "bold", size = 12, color = "#10231d"),
      legend.position = "none"
    )

  if (compare) {
    chart <- chart + facet_wrap(~pos_label, scales = "free_y", ncol = 2)
  }

  ggplotly(chart, tooltip = "text", height = if (compare) 600 else 430) |>
    layout(
      font = list(family = "DM Sans, sans-serif", color = "#17251f", size = 13),
      paper_bgcolor = "#ffffff",
      plot_bgcolor = "#ffffff",
      hoverlabel = list(
        bgcolor = "#10231d",
        bordercolor = "#c8f542",
        font = list(color = "#ffffff", family = "DM Sans, sans-serif")
      ),
      margin = list(l = 62, r = 24, t = 24, b = 62)
    ) |>
    config(displayModeBar = FALSE, responsive = TRUE)
}
```

::: {.summary-grid}
::: {.summary-card}
<span class="summary-label">QUALIFYING OBSERVATIONS</span>
<strong>`r format(n_story, big.mark = ",")`</strong>
<span class="summary-note">Players meeting the minutes threshold</span>
:::
::: {.summary-card}
<span class="summary-label">MINIMUM PLAYING TIME</span>
<strong>`r format(min(qualified_players$Min), big.mark = ",")` min</strong>
<span class="summary-note">Five full 90-minute matches</span>
:::
::: {.summary-card}
<span class="summary-label">LEAGUES COVERED</span>
<strong>`r league_count`</strong>
<span class="summary-note">European top-five competitions</span>
:::
:::

<details class="methodology">
<summary>Methodology</summary>
<p>Players qualify with non-missing age, goals, assists, and playing-time data, positive 90s, and at least 450 minutes played. Goal contributions per 90 are calculated as (goals + assists) / 90s. Age bands are 17–20, 21–24, 25–28, 29–32, and 33+; positions are grouped from the dataset's position codes. The overall chart shows individual-player distributions, while position comparisons use the mean within each age group. Each position panel in the comparison view has its own vertical scale.</p>
</details>

## Overall Distribution

Each box summarizes player-level goal contributions for one age group; the horizontal line is the median and the diamond is the mean. Hover over a point to see its metric and the age group's sample size.

::: {.chart-card}
### Goal Contributions per 90 by Age

```{r}
#| label: fig-overall-age-groups
#| fig-cap: "Boxplots show the distribution; the horizontal line marks the median and diamonds mark the mean."
#| warning: false
#| message: false
#| echo: false
age_colors <- c("#c8f542", "#9bcf43", "#62a94e", "#26744f", "#e87854")

overall_plot <- plot_ly(
  qualified_players,
  x = ~age_group,
  y = ~ga_per_90,
  type = "box",
  color = ~age_group,
  colors = age_colors,
  text = ~box_hover,
  hoverinfo = "text",
  boxpoints = "outliers",
  line = list(width = 1.5),
  marker = list(opacity = 0.65, size = 4),
  showlegend = FALSE
) |>
  add_trace(
    data = age_summary,
    x = ~age_group,
    y = ~mean_ga,
    type = "scatter",
    mode = "markers",
    marker = list(symbol = "diamond", size = 11, color = "#10231d", line = list(color = "#ffffff", width = 1.5)),
    text = ~paste0(
      "Age group: ", age_group,
      "<br>Mean: ", sprintf("%.4f", mean_ga), " G+A per 90",
      "<br>Sample size: ", sample_size
    ),
    hoverinfo = "text",
    showlegend = FALSE
  ) |>
  layout(
    xaxis = list(title = "Age group", categoryorder = "array", categoryarray = levels(qualified_players$age_group), tickfont = list(size = 12)),
    yaxis = list(title = "Goal contributions per 90", rangemode = "tozero", gridcolor = "#e1e8e2", tickfont = list(size = 12)),
    hovermode = "closest",
    font = list(family = "DM Sans, sans-serif", color = "#17251f", size = 13),
    plot_bgcolor = "#ffffff",
    paper_bgcolor = "#ffffff",
    margin = list(l = 72, r = 24, t = 24, b = 62),
    hoverlabel = list(bgcolor = "#10231d", bordercolor = "#c8f542", font = list(color = "#ffffff", family = "DM Sans, sans-serif"))
  ) |>
  config(displayModeBar = FALSE, responsive = TRUE)

overall_plot
```

<p class="chart-caption">Data: 2025–2026 season, top five European leagues. The box shows the spread of player observations; hover details include the age-band sample size.</p>
:::

## Compare by Position

Position changes the scale of goal involvement, so use the selector to focus on a role or compare all four roles side by side.

::: {.chart-card .position-card}
### Goal Contributions by Position

Bars show the mean goal contributions per 90 in each age group. Hover for the exact mean and group sample size.

<div class="position-control">
  <label for="position-select">Show</label>
  <select id="position-select" aria-controls="position-views">
    <option value="all">All positions</option>
    <option value="forwards">Forwards</option>
    <option value="midfielders">Midfielders</option>
    <option value="defenders">Defenders</option>
    <option value="goalkeepers">Goalkeepers</option>
  </select>
</div>

```{r}
#| label: position-comparison
#| warning: false
#| message: false
#| echo: false
make_position_description <- function(position) {
  peak <- position_summary |>
    filter(.data$pos_label == position) |>
    slice_max(.data$mean_ga, n = 1, with_ties = FALSE)

  paste0(
    position, " have their highest mean in the ", as.character(peak$age_group),
    " age group (", sprintf("%.4f", peak$mean_ga),
    " G+A per 90; n = ", peak$sample_size, ")."
  )
}

position_views <- tagList(
  tags$div(
    id = "position-view-all",
    class = "position-view",
    tags$p("Compare the four position groups. Each panel uses an independent vertical scale, so compare age patterns within a panel rather than bar heights between positions."),
    make_position_plot(position_summary, compare = TRUE)
  )
)

for (position in levels(position_summary$pos_label)) {
  position_data <- filter(position_summary, pos_label == position)
  position_id <- tolower(position)
  position_views <- tagAppendChild(
    position_views,
    tags$div(
      id = paste0("position-view-", position_id),
      class = "position-view",
      style = "display: none;",
      tags$p(make_position_description(position)),
      make_position_plot(position_data)
    )
  )
}

tags$div(id = "position-views", class = "position-chart", position_views)
```

<p class="chart-caption">Data: 2025–2026 season, top five European leagues (minimum 450 minutes). Labels and hover details use four decimal places; age groups are chronological. In “All positions,” panels use independent vertical scales.</p>

```{=html}
<script>
  (() => {
    const selector = document.getElementById('position-select');
    const views = document.querySelectorAll('#position-views .position-view');
    if (!selector || !views.length) return;

    function showPositionView() {
      const selected = selector.value;
      views.forEach((view) => {
        const active = view.id === `position-view-${selected}`;
        view.style.display = active ? '' : 'none';
        if (active) {
          view.querySelectorAll('.js-plotly-plot').forEach((plot) => {
            if (window.Plotly) window.Plotly.Plots.resize(plot);
          });
        }
      });
    }

    selector.addEventListener('change', showPositionView);
  })();
</script>
```
:::

</div>

## Standout Veterans

```{=html}
<section class="veteran-spotlight" aria-label="Lionel Messi standout veteran highlights">
  <div class="veteran-spotlight__media" aria-hidden="true">
    <iframe
      id="veteran-video"
      class="veteran-spotlight__video"
      src="https://www.youtube.com/embed/ONDoxgXEMYk?autoplay=1&mute=1&loop=1&playlist=ONDoxgXEMYk&controls=0&rel=0&modestbranding=1&playsinline=1&enablejsapi=1"
      title="Lionel Messi's best moments at the 2026 FIFA World Cup"
      allow="autoplay; encrypted-media; picture-in-picture"
      allowfullscreen
      loading="eager"
    ></iframe>
  </div>
  <div class="veteran-spotlight__overlay"></div>
  <div class="veteran-spotlight__content">
    <p class="veteran-spotlight__eyebrow">Standout veterans / Argentina</p>
    <h3>Lionel Messi</h3>
    <p class="veteran-spotlight__intro">A look at the moments, vision, and attacking craft that have kept one of soccer's defining players at the center of the game.</p>
    <div class="home-hero__actions">
      <button class="home-hero__button home-hero__button--secondary" id="veteran-video-toggle" type="button" aria-label="Pause background video" hidden>Pause</button>
      <a class="home-hero__button home-hero__button--primary" href="https://www.youtube.com/watch?v=ONDoxgXEMYk" target="_blank" rel="noopener">Watch highlights</a>
    </div>
    <p class="veteran-spotlight__note">2026 FIFA World Cup highlights. This feature is separate from the 2025–26 European-league dataset.</p>
  </div>
</section>

<script>
  (function() {
    const spotlight = document.querySelector('.veteran-spotlight');
    const frame = document.getElementById('veteran-video');
    const toggle = document.getElementById('veteran-video-toggle');
    if (!spotlight || !frame || !toggle) return;

    let player;
    let fallbackTimer;

    function showFallback() {
      window.clearTimeout(fallbackTimer);
      frame.src = 'https://www.youtube.com/embed/ONDoxgXEMYk?controls=1&rel=0&playsinline=1';
      spotlight.classList.add('veteran-spotlight--playing');
      toggle.hidden = true;
    }

    function onPlayerReady() {
      if (!window.YT || !window.YT.Player) {
        showFallback();
        return;
      }

      player = new window.YT.Player(frame, {
        events: {
          onReady: function(event) {
            event.target.mute();
            event.target.playVideo();
          },
          onStateChange: function(event) {
            if (event.data === window.YT.PlayerState.PLAYING) {
              window.clearTimeout(fallbackTimer);
              spotlight.classList.add('veteran-spotlight--playing');
              toggle.hidden = false;
              toggle.textContent = 'Pause';
              toggle.setAttribute('aria-label', 'Pause background video');
            } else if (event.data === window.YT.PlayerState.PAUSED) {
              toggle.textContent = 'Play';
              toggle.setAttribute('aria-label', 'Play background video');
            }
          },
          onError: showFallback
        }
      });
    }

    if (window.matchMedia('(prefers-reduced-motion: reduce)').matches) {
      showFallback();
      return;
    }

    fallbackTimer = window.setTimeout(showFallback, 12000);
    window.onYouTubeIframeAPIReady = onPlayerReady;

    toggle.addEventListener('click', function() {
      if (!player) return;
      if (player.getPlayerState() === window.YT.PlayerState.PLAYING) {
        player.pauseVideo();
      } else {
        player.playVideo();
      }
    });

    if (window.YT && window.YT.Player) {
      onPlayerReady();
    } else {
      const apiScript = document.createElement('script');
      apiScript.src = 'https://www.youtube.com/iframe_api';
      apiScript.onerror = showFallback;
      document.head.appendChild(apiScript);
    }
  })();
</script>
```

### Limitations & Causal Considerations

It is essential not to interpret these empirical patterns as proof that age directly causes changes in soccer performance:

1. **Selection & Survivor Bias:** Older players who continue earning regular playing time (450+ minutes) in Europe's top five leagues are elite survivors. We only observe the veterans good enough to stay in top leagues, not those whose performance declined and transferred to lower divisions.
2. **Measurement Limitations:** Goals and assists measure direct offensive end-product, but soccer performance involves press resistance, positioning, defensive recoveries, and stamina—attributes that may peak at different ages.
3. **Single-Season Cross-Section:** These data represent a single season snapshot (2025–2026) rather than a longitudinal cohort tracking individual athletes over their entire careers.

### Data Source
Data was retrieved from Kaggle: [Football Players Statistics (2025–2026)](https://www.kaggle.com/datasets/hubertsidorowicz/football-players-stats-2025-2026) by Hubert Sidorowicz, sourced from FBref.

